---
title: "Tidyverse_Start"
author: "Thema Monroe-White"
date: "2023-08-14"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

```

# R Script for tidyverse FOR STUDENTS

Read in the Palmer penguins dataset
```{r read penguins data, include=TRUE}
penguins <- read.csv("https://raw.githubusercontent.com/mhudsonvassell/coursefiles/main/penguins.csv")

```

Or go to File > Import dataset > from Text (base)

Load the tidyverse package (remember to install it first...)

To install, go to Tools > Install Packages > tidyverse
```{r load package, include=TRUE}
library("tidyverse")

```

Let's take a look at our data first
```{r view penguins, include=TRUE}
View(penguins)

```
Plot the penguins dataset, comparing 'bill_depth_mm' to 'bill_length_mm'


```{r compare bill depth and length, include=TRUE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(x = bill_depth_mm,
                  y = bill_length_mm)
  )
```

# CHALLENGE 1A
How has bill length changed over time? 
What do you observe if you plot length versus year?
Fill in for x and y

```{r bill length over time, include=TRUE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(x = year, 
                  y = bill_length_mm)
  )
```



# CHALLENGE 1B
Try a different geom_ function called geom_jitter. 
How is that different from geom_point?
Try a different plotting function

```{r use jitter, include=TRUE}
ggplot(data = penguins) +
  geom_jitter (
    mapping = aes(x = year, 
                  y = bill_length_mm)
  )
```



# Mapping data
See your plot points by category
Add color to the data points by 'island'
```{r color by island, include=TRUE}
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = bill_depth_mm, 
                  y = bill_length_mm, 
                  colour = island)
  )
```





# CHALLENGE 2
What will happen if you switch colour to also be by year? Is the graph still useful? 
Why or why not? What is the difference in the plot between when you colour by island and 
when you colour by year?
What do you add in for colour?

```{r color by various, include=TRUE}
ggplot(data = penguins) +
  geom_jitter(
    mapping = aes(x = bill_depth_mm, 
                  y = bill_length_mm,
                  colour =  year)
  )

```



There are other aesthetics we can use for the plot.
What if we want to change the size of the points?
We can add 'size' like we did for 'colour'
```{r size by various, include=TRUE}
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = bill_depth_mm, 
                  y = bill_length_mm, 
                  colour = species, 
                  size = species)
  )
```



Another option for aesthetics is shape, and we can combine things too
adjust the view so that both color and shape reflect the species of penguin
```{r shape by species, include=TRUE}
ggplot(data = penguins) + 
  geom_jitter(
    mapping = aes(x = bill_depth_mm, 
                  y = bill_length_mm, 
                  colour = species, 
                  shape = species)
  )
```



You may also just want to set one color for all your points
Let's set all the points to be blue
```{r single color, include=TRUE}
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = bill_depth_mm, 
                  y = bill_length_mm),
    colour = "blue"
  )
```




# CHALLENGE 4
Change the transparency (alpha) of the data points to 0.5.
Hint - this is another aesthetic feature like shape or colour
```{r set alpha, include=TRUE}
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = bill_depth_mm, 
                  y = bill_length_mm),
    alpha = 0.5
  )
```



# CHALLENGE 3
Change the transparency (alpha) of the data points by year
Hint - this is another aesthetic like shape, however, if it's by a variable, 
it needs to be placed within the mapping section of the code
```{r alpha by year, include=TRUE}
ggplot(data = penguins) + 
  geom_point(
    mapping = aes(x = bill_depth_mm, 
                  y = bill_length_mm, 
                  alpha = species
    )
  )
```




What if you want to add a regression line?
Plot bill depth (x-axis) by bill length (y-axis) like you did before 
and then layer on the regression using lm
Note: `geom_smooth()` using formula 'y ~ x'
```{r add regression line, include=TRUE}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, 
                     y = bill_length_mm)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")

```



# CHALLENGE 5
Modify the plot so the the points are coloured by island, 
The transparency remains .5
and there is a single regression line.
```{r combine aesthetics, include=TRUE}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, 
                     y = bill_length_mm)) +
  geom_point(mapping = aes(colour = island),
             alpha = .5) +
  geom_smooth(method = "lm")
```


  

# CHALLENGE 6
We may want to separate the visualizations of each species by sex. 
We can do that with faceting.
```{r facet by sex, include=TRUE}
ggplot(penguins, 
       aes(x = bill_depth_mm, 
           y = bill_length_mm,
           colour = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~sex)
```

